---
title: "1 · From awards to outputs"
subtitle: "RePORTER awards → publications (with iCite RCR) → clinical trials"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
# Native-R connectors (no Python). root.dir = project root so the cache resolves.
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(tibble)
  for (f in c("cache.R", "reporter.R", "icite.R", "clinicaltrials.R"))
    source(file.path(proj, "R", f))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This notebook walks the core spine of the platform in **pure R** (via `httr2` /
`jsonlite`): find NIH **awards**, follow their authoritative **publication
links**, enrich with **iCite** citation metrics, and surface the **clinical
trials** an award funded. Responses cache under `data/cache/`, so re-running is
fast and offline after the first pass.

## Find awards

```{r}
awards <- reporter_search_projects(
  query = "PFAS proteomics", institutes = "NIEHS",
  fiscal_years = 2018:2024, limit = 5
)
reporter_as_tibble(awards) |> select(core_project_num, fiscal_year, ic, total_cost, title)
```

### Search by investigator

```{r}
khatri <- reporter_search_projects(pi_names = "Purvesh Khatri", limit = 100)
cat(length(khatri), "awards; institutions:\n")
unique(vapply(khatri, function(a) a$organization %||% NA_character_, character(1)))
```

## Link an award to its publications

The authoritative publication links from RePORTER, enriched with iCite's
field-normalized **Relative Citation Ratio** (1.0 = median NIH-funded paper).

```{r}
pmids <- reporter_get_publications("R01ES032470", limit = 30)
icite_fetch_metrics(pmids) |>
  arrange(desc(rcr)) |>
  transmute(pmid, year, rcr = round(rcr, 1), citations, title = substr(title, 1, 60)) |>
  head(8)
```

## Citation metrics directly

```{r}
icite_fetch_metrics(c("22745249", "36656942")) |>   # CRISPR landmarks
  select(pmid, year, rcr, citations, nih_percentile)
```

## Clinical trials an award funded

Trials that *self-report* the grant number give an authoritative award→trial link.

```{r}
trials <- ct_find_trials_for_grant("R01DK075877")
tibble(
  nct_id = vapply(trials, \(t) t$nct_id, character(1)),
  status = vapply(trials, \(t) t$overall_status %||% NA, character(1)),
  phase  = vapply(trials, \(t) t$phase %||% NA, character(1)),
  title  = vapply(trials, \(t) substr(t$brief_title, 1, 60), character(1))
)
```
