---
title: "2 · Knowledge graph & linkage coverage"
subtitle: "Award→publication edges, and where they are missing"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(ggplot2)
  for (f in c("cache.R", "reporter.R", "duckdb_store.R"))
    source(file.path(proj, "R", f))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The platform's central object is a **knowledge graph**: awards linked to PIs,
institutions, ICs, FOAs, publications, and trials — every edge carrying
provenance. The single most important property of that graph is its **coverage**:
*which awards actually have the edges an analysis assumes?* Coverage is never
uniform, and the design doc's rule is to disclose the gradient before analyzing
on top of it.

## One award's neighborhood (a single edge list)

A grant's authoritative publication edges come straight from RePORTER.

```{r}
proj_award <- reporter_search_projects(query = "single cell", fiscal_years = 2015, limit = 1)[[1]]
core <- proj_award$core_project_num
pmids <- reporter_get_publications(core, limit = 50)
cat(sprintf("award %s (%s, FY%s)\n  %d linked publications: %s ...\n",
            core, proj_award$ic, proj_award$fiscal_year, length(pmids),
            paste(head(pmids, 6), collapse = ", ")))
```

Each edge is `award --produced--> publication`, asserted by NIH itself (not a
text-match guess). That provenance is what lets coverage be measured honestly.

## Coverage at scale: who has the edges?

The bulk DuckDB store holds every core project and its publication links, so we
can compute the fraction of grants with ≥1 linked publication and **stratify** it.

```{r}
con <- bulk_connect()
```

### By Institute / Center (FY2010)

```{r cov-ic, fig.width=6.5, fig.height=3.6}
cov_ic <- bulk_coverage_stratified(con, by = "ic", years = 2010, min_grants = 300)
cov_ic |> head(15)
cov_ic |> head(15) |>
  ggplot(aes(reorder(strat, coverage_pct), coverage_pct)) +
  geom_col(fill = "steelblue") + coord_flip() +
  labs(x = "IC", y = "% of grants with >= 1 linked pub",
       title = "Publication-link coverage by IC (FY2010)") + theme_minimal()
```

### By mechanism family (NCI)

```{r}
bulk_coverage_stratified(con, by = "family", ic = "NCI", years = 2008:2012)
```

Coverage is high for research and center grants but the **`other`** bucket
(contracts, non-research codes) has near-zero publication edges — a real gap, not
noise. Any productivity or translation analysis must exclude or flag it; pretending
those grants simply produced nothing would be a measurement artifact.

### By grant age (the censoring gradient)

```{r cov-age, fig.width=6.5, fig.height=3.2}
cov_age <- bulk_coverage_stratified(con, by = "fiscal_year", ic = "NCI",
                                    years = 2005:2018, min_grants = 100)
cov_age |> arrange(strat) |>
  ggplot(aes(strat, coverage_pct)) + geom_line(color = "firebrick") +
  geom_point(color = "firebrick") +
  labs(x = "fiscal year", y = "% with >= 1 linked pub",
       title = "Coverage falls for recent grants (right-censoring)") + theme_minimal()
```

Recent fiscal years show **lower** coverage — not because new grants are
unproductive, but because publications and their RePORTER links lag the award by
years. This right-censoring is exactly why every downstream analysis reports the
window and the coverage alongside the result.

```{r}
bulk_disconnect(con)
```

> **The discipline.** Coverage is a first-class output here, not an afterthought.
> An edge that doesn't exist is reported as missing — never silently treated as a
> zero — so the graph's gaps shape the conclusions instead of hiding inside them.
