---
title: "3 · Research productivity (publications per dollar)"
subtitle: "Output-per-$ at population scale — split by mechanism family"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(ggplot2)
  source(file.path(proj, "R", "cache.R")); source(file.path(proj, "R", "duckdb_store.R"))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

"Publications per dollar" is only honest **within a mechanism family**. A shared
center core (`U54`, `P30`) carries dozens of papers under one award; a training
grant (`T32`, `F31`) carries few by design. Ranking them on one axis is a
category error — so we classify every core project by activity code first.

> **Prerequisite (one time):** the Python `nih-agent bulk build` / `build-pubs`
> populated `data/processed/nih.duckdb` (2.1M awards, 6.1M publication links).
> R opens that same file directly.

```{r}
con <- bulk_connect()
```

## Output per $1M by mechanism family (NCI, FY2010)

```{r}
fam <- bulk_productivity_by_mechanism(con, ic = "NCI", years = 2010)
fam
```

```{r fam-plot, fig.width=6.5, fig.height=3.2}
fam |> filter(family != "other") |>
  ggplot(aes(reorder(family, median_pubs_per_M), median_pubs_per_M)) +
  geom_col(fill = "steelblue") + coord_flip() +
  labs(x = NULL, y = "median linked pubs per $1M",
       title = "NCI FY2010 — output per dollar by mechanism family") +
  theme_minimal()
```

Two things the split makes visible and a single ranking would hide:

- **Center cores** post the most publications per dollar (they aggregate the
  output of many sub-projects under one administrative award).
- The **`other`** bucket — contracts and non-research activity codes — links to
  essentially **zero** publications. Folding it into a research ranking would
  silently depress the denominator. Naming it is the honest move.

## The same split is portfolio-wide, not NCI-specific

```{r}
bind_rows(
  cbind(ic = "NIAID", bulk_productivity_by_mechanism(con, ic = "NIAID", years = 2010) |>
          filter(family == "research") |> select(median_pubs_per_M, grants)),
  cbind(ic = "NIGMS", bulk_productivity_by_mechanism(con, ic = "NIGMS", years = 2010) |>
          filter(family == "research") |> select(median_pubs_per_M, grants)),
  cbind(ic = "NHLBI", bulk_productivity_by_mechanism(con, ic = "NHLBI", years = 2010) |>
          filter(family == "research") |> select(median_pubs_per_M, grants))
)
```

```{r}
bulk_disconnect(con)
```

> **Caveats.** Publication links are authoritative (RePORTER's own attribution)
> but right-censored for recent years; per-dollar output is a *proxy* for
> productivity, not for value; and the median, not the mean, is reported because
> a handful of mega-cores skew the average upward.
