---
title: "4 · Health pulse"
subtitle: "Juxtaposing NIH inputs against the nation's outcome for one condition"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(ggplot2)
  for (f in c("cache.R", "reporter.R", "clinicaltrials.R", "cdc.R", "fda.R", "conditions.R"))
    source(file.path(proj, "R", f))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

For one canonical condition we assemble the full arc on a single page:

```
Input    : NIH funding by fiscal year, award count   (RePORTER)
Output   : clinical trials                            (ClinicalTrials.gov)
Translate: FDA-approved drugs for the indication      (openFDA)
Outcome  : age-adjusted mortality trend               (CDC NCHS)
```

The crosswalk is a small **curated, auditable dictionary** — not an ML matcher.
Each condition carries the handles needed to pull every series.

```{r}
list_conditions()
```

## A pulse for diabetes (FY2008–2017)

```{r pulse}
p <- condition_pulse("diabetes", years = 2008:2017, per_year_cap = 1000)
cat(sprintf("%s · $%.2fB over %d-%d · %d distinct awards%s\n",
            p$condition_label, p$total_funding / 1e9,
            min(p$years), max(p$years), p$distinct_awards,
            if (p$funding_is_floor) " (a FLOOR — some years hit the record cap)" else ""))
cat(sprintf("Trials: %d (%d with posted results) · FDA-labeled drugs: %d\n",
            p$trial_count, p$trials_with_results, p$approved_drug_count))
```

### Input: NIH funding trend

```{r funding, fig.width=6.5, fig.height=3.2}
ggplot(p$funding_by_year, aes(fiscal_year, total_funding / 1e9)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = p$years) +
  labs(x = NULL, y = "NIH funding ($B)",
       title = sprintf("%s — NIH funding by fiscal year", p$condition_label)) +
  theme_minimal()
```

### Outcome: age-adjusted mortality

```{r mortality, fig.width=6.5, fig.height=3.2}
p$mortality
ggplot(p$mortality, aes(year, aadr)) +
  geom_line(color = "firebrick") + geom_point(color = "firebrick") +
  labs(x = NULL, y = "age-adjusted deaths / 100k",
       title = sprintf("%s — US age-adjusted mortality", p$condition_label)) +
  theme_minimal()
```

```{r}
cat(sprintf("Mortality (AADR) change across the window: %+.1f%%\n",
            p$mortality_aadr_change_pct))
```

### Translation: FDA-labeled drugs

```{r}
cat(p$approved_drug_count, "labels match the indication. Examples:\n")
cat(paste("  -", p$example_drugs), sep = "\n")
```

## Why this is a juxtaposition, not a scorecard

```{r}
cat(strwrap(p$caveat, width = 86), sep = "\n")
```

Funding rose ~13% across the decade while age-adjusted diabetes mortality was
roughly flat (`r sprintf("%+.1f%%", p$mortality_aadr_change_pct)`). That is **not**
an indictment of NIH: mortality reflects a 10–30y research lag, countless non-NIH
drivers, and reactive funding (dollars often rise *because* a problem is worsening).
The platform's stance is to **show the series side by side** and refuse to compute
a "return per death averted." The honest read is the contrast itself — and the
disclosed caveats that travel with it.
