---
title: "5 · Diminishing returns"
subtitle: "Does research output per dollar fall as a PI's grant support rises?"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(ggplot2)
  for (f in c("cache.R", "reporter.R", "icite.R", "clinicaltrials.R", "analysis.R"))
    source(file.path(proj, "R", f))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The NIH Extramural Nexus ("Open Mike") repeatedly found **diminishing returns**
to grant support: as a PI's total funding rises, research **output per dollar**
falls. Here it is reproduced in pure R — support = total RePORTER funding per PI,
output = summed iCite **Relative Citation Ratio (RCR)** over their linked papers.

```{r}
dr <- diminishing_returns(institutes = "NIDDK", query = "beta cell",
                          fiscal_years = 2014:2015, max_pis = 24)
bins <- dr |>
  group_by(quartile) |>
  summarise(n_pis = n(), mean_funding = mean(funding),
            mean_rcr_per_million = mean(rcr_per_million), .groups = "drop")
bins
```

Output-per-dollar declines across the lower quartiles — the diminishing-returns
signature (the top quartile holds far more money for proportionally less output).

```{r dr-plot, fig.width=6, fig.height=3.5}
ggplot(bins, aes(factor(quartile), mean_rcr_per_million)) +
  geom_col(fill = "steelblue") +
  labs(x = "PI funding quartile", y = "RCR per $1M (output per dollar)",
       title = "Diminishing returns: output per dollar by funding quartile") +
  theme_minimal()
```

> **Caveats.** Output is lifetime summed RCR over authoritative pub links (recent
> awards are right-censored); support is a funding proxy for NIH's Grant Support
> Index; this is a bounded, single-topic sample. Compare *within* a cohort.
