---
title: "6 · Do RFAs produce different science than parent PAs?"
subtitle: "A confounded raw comparison, corrected with a controlled regression"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr)
  source(file.path(proj, "R", "cache.R")); source(file.path(proj, "R", "duckdb_store.R"))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

Does the *type* of funding opportunity relate to output? The **raw** comparison
is confounded — RFAs fund bigger grants and bigger mechanisms. Here it is
corrected with a controlled regression over the full ExPORTER bulk data.

```{r}
con <- bulk_connect()
df <- bulk_grant_panel(con, ic = "NCI", years = 2006:2010, funding_floor = 1e5)
bulk_disconnect(con)

df <- df |> filter(foa_type %in% c("RFA", "PA", "PAR", "NONE"))
keep <- df |> count(activity_code) |> filter(n >= 30) |> pull(activity_code)
df <- df |> filter(activity_code %in% keep) |>
  mutate(log_funding = log(funding), log_pubs = log1p(pubs),
         foa_type = relevel(factor(foa_type), ref = "PA"))
sprintf("%d grants · %d mechanisms · NCI FY2006–2010", nrow(df), length(keep))
```

## The naïve picture (confounded)

```{r}
df |> group_by(foa_type) |>
  summarise(grants = n(), mean_funding = round(mean(funding)),
            mean_pubs = round(mean(pubs), 1), .groups = "drop")
```

Taken at face value, **RFA grants look the most productive**. But they also fund
larger grants and different mechanisms.

## The controlled regression

`log(1 + publications)` on FOA type, holding **mechanism, fiscal year, and (log)
funding** constant (PA is the reference).

```{r}
m <- lm(log_pubs ~ foa_type + log_funding + factor(activity_code) + factor(fiscal_year),
        data = df)
co <- summary(m)$coefficients
fo <- co[grepl("^foa_type", rownames(co)), , drop = FALSE]
tibble(vs_PA = sub("foa_type", "", rownames(fo)),
       pct_diff = round((exp(fo[, "Estimate"]) - 1) * 100),
       p = signif(fo[, "Pr(>|t|)"], 2))
cat(sprintf("R^2 = %.2f, n = %d\n", summary(m)$r.squared, nrow(df)))
```

The sign **flips**: adjusted for mechanism, year, and funding, **RFA-funded grants
produce ~30–35% *fewer* publications than parent-PA grants**. The raw advantage was
entirely a confound.

> **Caveats.** Publications are a narrow output (RFAs target cohorts, datasets,
> trials, translation that pubs miss); the model controls for mechanism/funding/
> year but not topic, PI track record, or grant intent; linkage is
> authoritative-only. It says *what*, not *why* — not "RFAs are wasteful."
