---
title: "7 · Discovery → translation"
subtitle: "Tracing a basic-science grant's reach into the clinic via citations"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr)
  for (f in c("cache.R", "reporter.R", "icite.R", "clinicaltrials.R", "analysis.R"))
    source(file.path(proj, "R", f))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The direct award→trial link misses the basic grant that *enabled* a therapy
years later. We recover that chain through the **citation graph**, using a signal
NIH precomputes: iCite's `cited_by_clin` — the *clinical* articles citing a paper.

```
award → publications → clinical citers (iCite) → distinct clinical reach + APT
```

## Ranking a portfolio: which GWAS grants reached the clinic?

```{r}
gwas <- translation_scan(query = "GWAS", fiscal_years = 2008:2014, max_grants = 12)
gwas |> head(8)
```

The leaders are pharmacogenomics (anti-TNF response, drug transporters) — and
their high **APT** (Approximate Potential to Translate, ~0.7) corroborates that
they are genuinely translational, not just well-cited basic science.

## How do different technologies reach the clinic?

```{r}
summarise_tech <- function(q, yrs) {
  s <- translation_scan(query = q, fiscal_years = yrs, max_grants = 12)
  tibble(top_reach = max(s$clinical_reach), median_reach = median(s$clinical_reach),
         mean_APT = round(mean(s$mean_apt, na.rm = TRUE), 2))
}
bind_rows(
  cbind(technology = "GWAS",        summarise_tech("GWAS", 2008:2014)),
  cbind(technology = "Single-cell", summarise_tech("single cell sequencing", 2012:2017)),
  cbind(technology = "Exposomics",  summarise_tech("exposome", 2012:2019))
)
```

GWAS reaches the clinic most (pharmacogenomics); single-cell is younger; and
**exposomics' reach runs through infrastructure** (centers/cohorts), not a drug
pipeline. The platform makes that contrast characterizable from public data.

> **Caveats.** Clinical reach is *inferred* (papers iCite flags clinical that cite
> the grant's work), not therapies; centers/training grants inflate reach (APT
> helps separate); younger fields are right-censored.
