---
title: "8 · Population-scale analyses (ExPORTER → DuckDB)"
subtitle: "Funding concentration, output-per-$, FOA type, and latency — no API caps"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
proj <- normalizePath("..")
knitr::opts_knit$set(root.dir = proj)
suppressMessages({
  library(dplyr); library(ggplot2)
  source(file.path(proj, "R", "cache.R")); source(file.path(proj, "R", "duckdb_store.R"))
})
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

The RePORTER API caps result offsets below NIH's annual award count, so all-NIH
questions need the **bulk ExPORTER files in DuckDB**. R opens the same store the
Python `bulk build` produced (2.1M awards, 6.1M publication links).

> **Prerequisite (one time):** `nih-agent bulk build 2000:2025` and
> `nih-agent bulk build-pubs 2000:2025` (the Python CLI builds the DuckDB file).

```{r}
con <- bulk_connect()
```

## Funding concentration (top-decile share + Gini)

```{r}
cc <- bulk_concentration(con, ic = "NCI", year = 2021, top_n = 5)
cat(sprintf("NCI FY2021: %d PIs · $%.1fB · top 1%%=%.0f%% · top 10%%=%.0f%% · Gini %.2f\n",
            cc$n_pis, cc$total / 1e9, 100 * cc$top1, 100 * cc$top10, cc$gini))
cc$top |> transmute(pi_name, funding = round(funding), awards)
```

> The extreme top is inflated by large contracts/center grants under one
> administrative contact PI — aggregate shares and Gini are the robust read.

## Diminishing returns at full scale

```{r dr, fig.width=6, fig.height=3.5}
p <- bulk_funding_vs_output(con, ic = "NCI", year = 2010, funding_floor = 250000)
bins <- p |> group_by(quartile) |>
  summarise(n = n(), mean_funding = mean(funding),
            mean_pubs = mean(pubs), pubs_per_M = mean(pubs_per_million), .groups = "drop")
bins
ggplot(bins, aes(factor(quartile), pubs_per_M)) + geom_col(fill = "steelblue") +
  labs(x = "PI funding quartile", y = "linked pubs per $1M",
       title = "NCI FY2010 — output per dollar by funding quartile") + theme_minimal()
```

## Output by FOA type (RFA vs PA)

```{r}
bulk_foa_types(con, ic = "NCI", year = 2010) |>
  transmute(foa_type, awards, funding_M = round(funding / 1e6),
            pubs, pubs_per_million = round(pubs_per_million, 1))
```

## Grant → first-publication latency (with censoring)

```{r}
lat <- bulk_latency(con, years = 2005:2012)
cat(sprintf("%d grants · %d with a linked pub\n  median %.0fy (IQR %.0f–%.0f) · %.0f%% censored\n",
            lat$grants_total, lat$grants_with_output, lat$median, lat$p25, lat$p75,
            100 * lat$pct_censored))
```

```{r}
bulk_disconnect(con)
```

Censoring is reported, not hidden: silently dropping the ~40% of grants with no
linked publication would bias the median downward.
